---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%",
  dpi = 330,
  message = FALSE, warning = FALSE
)

```
# Baseline nowcast 

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/epinowcast/primarycensored/workflows/R-CMD-check/badge.svg)](https://github.com/epinowcast/baselinenowcast/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/epinowcast/baselinenowcast/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epinowcast/baselinenowcast)

[![Universe](https://epinowcast.r-universe.dev/badges/primarycensored)](https://epinowcast.r-universe.dev/baselinenowcast)
[![MIT
license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/epinowcast/baselinenowcast/blob/master/LICENSE.md/)
[![GitHub contributors](https://img.shields.io/github/contributors/epinowcast/baselinenowcast)](https://github.com/epinowcast/baselinenowcast/graphs/contributors)
<!-- badges: end -->

## Summary

```{r, results = "asis", echo=FALSE}
cat(gsub("\n[ ]+", " ", packageDescription("baselinenowcast")$Description))
```

## Installation

<details><summary>Installing the package</summary>

```{r, child="vignettes/chunks/_readme-install-baselinenowcast.Rmd"}
```

</details>

## Resources

We provide a range of other documentation, case studies, and community spaces to ask (and answer!) questions:

<details><summary>Organisation Website</summary>

Our [organisation website](https://www.epinowcast.org/) includes links to other resources, [guest posts](https://www.epinowcast.org/blog.html), and [seminar schedule](https://www.epinowcast.org/seminars.html) for both upcoming and past recordings.

</details>

<details><summary>Community Forum</summary>

Our [community forum](https://community.epinowcast.org/) has areas for [question and answer](https://community.epinowcast.org/c/interface/15) and [considering new methods and tools](https://community.epinowcast.org/c/projects/11), among others. If you are generally interested in real-time analysis of infectious disease, you may find this useful even if you do not use `baselinenowcast`.

</details>

## Contributing

We welcome contributions and new contributors! We particularly appreciate help on [identifying and identified issues](https://github.com/epinowcast/baselinenowcast/issues). Please check and add to the issues, and/or add a [pull request](https://github.com/epinowcast/baselinenowcast/pulls) and see our [contributing guide](https://github.com/epinowcast/.github/blob/main/CONTRIBUTING.md) for more information.


### How to make a bug report or feature request

Please briefly describe your problem and what output you expect in an [issue](https://github.com/epinowcast/baselinenowcast/issues). See our [contributing guide](https://github.com/epinowcast/.github/blob/main/CONTRIBUTING.md) for more information.

### Code of Conduct

Please note that the `baselinenowcast` project is released with a [Contributor Code of Conduct](https://github.com/epinowcast/.github/blob/main/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## Citation

If you use `baselinenowcast` in your work, please consider citing it with `citation("baselinenowcast")`.

## Contributors

<!-- will add this in later -->



