
---
title: "Getting Started with baselinenowcast"
description: "A quick start example demonstrating use of baselinenowcast"
author: Kaitlyn Johnson
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting Started with baselinenowcast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Incomplete reporting of epidemiological data at recent times can result in case
count data that is right-truncated.
Right-truncated case counts can be misleading to interpret at face-value, as
they will typically show a decline in the number of reported observations in
the most recent time points.
These are the time points where the highest proportion of the data has yet to
be observed in the dataset.

The imputation of the cases that will eventually be observed up until the
current time is referred to as a nowcast.

A number of methods have been developed to nowcast epidemiological case count
data.

The purpose of `baselinenowcast` is to provide a nowcast computed directly from
the most recent observations to estimate a delay distribution empirically, and
apply that to the partially observed data to generate a nowcast.

In the below section, we will describe an example of a nowcasting problem, and
demonstrate how to use `baselinenowcast` to estimate a delay distribution from
the data and apply that estimate to generate a probabilistic nowcast.

# Packages

As well as the `baselinenowcast` package this vignette also uses `epinowcast`,
`ggplot2`, and `dplyr`.
The installation of `epinowcast` is not required for using the package,
however, its pre and post-processing functions provide a lot of the data
wrangling needed to set up the nowcasting problem.
We note that no components of the vignette require installing `CmdStan`,
which is a downstream dependency of `epinowcast`.
We will just be using the `R` components of `epinowcast`, which can be
installed using the example lines of code below, so there is no need to
additionally install `CmdStan`.
```{r setup, message = FALSE}
# Installing epinowcast
# install.packages( #nolint
#  "epinowcast", repos = "https://epinowcast.r-universe.dev" #nolint
# ) #nolint
# Load packages
library(baselinenowcast)
library(epinowcast)
library(ggplot2)
library(dplyr)
# Set seed for reproducibility
set.seed(123)
```

# Data

Nowcasting of right-truncated case counts involves the estimation of reporting
delays for recently reported data.
For this, we need case counts both by when they were diagnosed (often
called "reference date") and by when they were
reported (i.e. when administratively recorded via public health surveillance;
often called "report date"). The difference between the reference date and the
report date is the reporting delay.
For this quick start, we use daily level data from the
[Robert Koch Institute via the Germany Nowcasting hub](https://github.com/KITmetricslab/hospitalization-nowcast-hub/wiki/Truth-data#role-an-definition-of-the-seven-day-hospitalization-incidence).
These data represent hospitalisation counts by date of positive test and date
of test report in Germany up to October 1, 2021.


# Filtering and plotting the data

We will filter the data to just look at the national-level data, for all age
groups.
We will pretend that we are making a nowcast as of July 1, 2021, therefore we
will exclude all reference dates and report dates after that date.
`germany_covid19_hosp` is provided as package data from `epinowcast`
Let's start by plotting the sum of the reports at each reference date,
and then compare that to what we will eventually observe as of the final date
in the complete dataset.
The red line shows the cumulative number of confirmed admissions on each report
date, across all delays, using the data available as of July 1, 2021.
It demonstrates the characteristic behavior of right-truncation.
This is because we have not yet observed the data that will become available
for the longer delays at recent time points.

Our task will be to estimate what the "final" cumulative number of cases will
at each reference date, observed as of the "fully observed" data on October
2021.
```{r plot-the-data-by-reference-date}
data_long <- germany_covid19_hosp # import data from epinowcast

ggplot() +
  geom_line(
    # Plot the data summed across reporting days
    data = data_long |>
      enw_filter_report_dates(latest_date = "2021-07-01") |>
      filter(
        location == "DE", age_group == "00+",
        report_date == "2021-07-01"
      ),
    aes(x = reference_date, y = confirm), color = "darkred"
  ) +
  geom_line(
    data = data_long |>
      filter(
        location == "DE", age_group == "00+",
        reference_date <= "2021-07-01"
      ) |>
      group_by(reference_date) |>
      summarise(confirm = max(confirm)),
    aes(x = reference_date, y = confirm), color = "black"
  ) +
  theme_bw() +
  xlab("Reference date") +
  ylab("Confirmed admissions") +
  scale_y_continuous(trans = "log10") +
  ggtitle("Comparing real-time and later observed cases")
```
Here the black line represents the quantity we will evaluate our nowcast
against, and the red line represents the data we have available to us up until
July 1st, 2021.

# Pre-processing

In order to compute a nowcast for this data, we will need to start by creating
what we call a reporting triangle.
This is a wide formatted dataframe where each row represents one of the time
points being reference and each column represents the delay, starting from 0 up
until the maximum delay.
The entries represent the number of new cases assigned to that reference
time point with a particular delay.
The reporting triangle will be used to estimate the delay distribution, or the
proportion of the final number of cases reported on a particular delay.
Since this data is both reported and referenced daily, we will use the time
scale of days to create the reporting triangle, but the delay and the reference
date can have any temporal granularity.

In this example, we will both fit our delay distribution, and apply it to
generate a nowcast using the same data, the national level data from Germany
for all age groups.
However, these components can be separated, so for example, we could use the
national level data for all age groups to estimate a delay distribution, and
then we could apply that elsewhere, for example to the data stratified by
age group and location.
This type of "borrowing" from another training dataset can be really useful
when you have low counts or relatively sparse data, which is likely to be the
case for smaller populations.

In the below sections, we will specify our nowcast date, the maximum delay,
and the number of observations by reference date that we want to use to
estimate the delay distribution.
We recommend choosing the maximum delay and number of historical observations
based on an exploratory data analysis, as these specifications will change
significantly depending on the dataset.
```{r user-specificatons}
nowcast_date <- "2021-07-01"
# Specify the maximum delay, which will determine the length of your delay
# distribution. Empirical data outside this delay window will not be used for
# training.
max_delay <- 40
# Specify the number of reference dates to use to estimate the delay
# distribution.Note this assumes you want the most recent observations
# (though we can consider changing this)
n_history <- 60
```

Next we will use the `epinowcast` function, `enw_preprocess_data()` and the
data in the form a of a long tidy dataframe indexed by reference date and
report date and filtered to the strata we are interested in, to generate a
reporting triangle.
```{r use-epinowcast-preprocessing}
# Noting that this is the only way epinowcast preprocessing would work --
# return to this later. IDate was throwing errors if we used the dplyr processed
# observed long above.
observed_long <- data_long[location == "DE"][age_group == "00+"] |> # nolint
  enw_filter_report_dates(latest_date = nowcast_date) |>
  enw_filter_reference_dates(include_days = n_history - 1)
head(observed_long)
# Get the reporting triangle, adding an additional day because epinowcast
# we want the max_delay + 1 entries since 0 is a valid delay.
pobs <- enw_preprocess_data(
  obs = observed_long,
  max_delay = max_delay + 1
)

triangle_full <- pobs$reporting_triangle[[1]]
head(triangle_full)

# Check that the sum of the rows in the reporting triangle is the
# same as the sums across report dates by reference date
observed_sums <- observed_long |>
  group_by(reference_date) |>
  summarise(total_confirmed = max(confirm)) |>
  mutate(triangle_sums = rowSums(triangle_full |>
    select(-`.group`, -reference_date) |>
    as.matrix() |>
    unname()))
```

# Estimate delay

Now that we have the reporting triangle, we are now ready to pass it in to the
`baselinenowcast` package to estimate the delay distribution.
We will tell the function the maximum delay and the number of observations we
want to use, though the default will be to use the whole reporting triangle.
If the reporting triangle is too small for the user-specified delays and number
of training observations, the function will throw an error.
We only want to pass in the reporting triangle (for a single group!) to this
function.
If reference date are repeated because the reporting triangle contains multiple
strata, the function will throw an error.

The `get_delay_estimate()` function expects the following inputs:
  - `triangle`: a matrix with the reporting triangle for a single strata. Here
  the rows represent the time points and the columns represent the observed
  delays, starting at 0.
  - `max_delay`: an integer indicating the maximum delay to estimate. This must
  be less than or equal to the number of rows in `triangle` minus 1, since we
  assume `triangle` is indexed at 0.
  - `n_history`: an integer indicating the number of observations by reference
  date to use to fit the delay distribution. This must be less than or equal to
  the number of rows in `triangle`.
`enw_preprocess()` returns a `triangle` with the columns `.group` and
`.reference_date` and delays indicated by the column names. While we will
eventually write methods that will map from other input formats such as this,
we will start by demonstrating the functionality on only the matrix of the
reporting triangle.

```{r estimate-delay}
triangle <- triangle_full |>
  select(-`.group`, -reference_date) |> # remove unnecessary columns
  as.matrix() |>
  unname()
head(triangle)

delay_pmf <- get_delay_estimate(
  triangle = triangle,
  max_delay = max_delay,
  n_history = n_history
)

delay_df <- data.frame(
  delay = 0:(length(delay_pmf) - 1),
  pmf = delay_pmf
)

ggplot(delay_df) +
  geom_line(aes(x = delay, y = cumsum(pmf))) +
  xlab("Delay") +
  ylab("Cumulative proportion reported") +
  ggtitle("Empirical point estimate of cumulative proportion reported by delay") + # nolint
  theme_bw()

ggplot(delay_df) +
  geom_line(aes(x = delay, y = pmf)) +
  xlab("Delay") +
  ylab("Proportion reported") +
  ggtitle("Empirical point estimate of proportion reported by delay") +
  theme_bw()
```

# Apply delay to generate point nowcast

The next step in our workflow is to take the estimated delay distribution and
apply it to the partially observed reporting triangle, generating an estimate
of the number of new cases confirmed at each reference date and delay.
This will generate a point estimate of what we can call the reporting square,
which is the complete set of reference dates and delays.
In this case, we will be applying the delay to the same reporting triangle we
used to generate the estimate, but this doesn't always have to be the case.
The reporting triangle we are applying it to must have the same `max_delay`
as the delay estimate.

```{r}
point_reporting_square <- apply_delay(
  triangle_to_nowcast = triangle,
  delay_pmf = delay_pmf
)
```

We'll make a quick plot to compare the nowcasted confirmed cases through
July 1, 2021, from the observations up until October 1, 2021.
We'll compare this to the right-truncated data available up until July 1, 2021.
```{r}
final_data <- data_long[location == "DE"][age_group == "00+"] |> # nolint
  enw_filter_report_dates(latest_date = "2021-10-01") |>
  enw_filter_reference_dates(
    latest_date = "2021-07-01",
    include_days = n_history - 1
  ) |>
  group_by(reference_date) |>
  summarise(
    total_confirmed = max(confirm)
  ) |>
  mutate(nowcast = rowSums(point_reporting_square))

ggplot() +
  geom_line(
    # Plot the data summed across reporting days as of July 1,2021
    data = observed_long |>
      group_by(reference_date) |>
      summarise(total_confirmed = max(confirm)),
    aes(x = reference_date, y = total_confirmed), color = "darkred"
  ) +
  geom_line(
    data = final_data,
    aes(x = reference_date, y = total_confirmed), color = "black"
  ) +
  geom_line(
    data = final_data,
    aes(x = reference_date, y = nowcast), color = "darkblue"
  ) +
  theme_bw() +
  xlab("Reference date") +
  ylab("Confirmed admissions") +
  scale_y_continuous(trans = "log10") +
  ggtitle("Comparing real-time, nowcasted, and later observed cases")
```

Here we can see that our point nowcast slightly overestimates what was
eventually reported (black line), but does a decent overall job of correcting
for the right-truncation observed in the red line (the data prior to the
nowcast).

What we really want is a probabilistic estimate of the nowcasted
case counts.


# Estimate uncertainty

So far, we've demonstrated how to generate a point estimate of a nowcast.
We would like to generate probabilistic nowcasts, which will require
estimating the empirical error between the expected observations and the
actual observations when applying the delay distribution.
This assumes a negative binomial observation model.

We will start by estimating the dispersion parameter of the negative binomial
as a function of the delay d


```{r}
disp_params <- estimate_uncertainty(
  triangle_to_nowcast = triangle,
  delay_pmf = delay_pmf
)
```

# Generate probabilistic nowcast

Now that we have estimated the dispersion, we can use the delay distribution,
the reporting triangle that we want to nowcast, and the dispersion parameters to
generate a probabilistic nowcast.

```{r}
# nowcast <- generate_nowcast( #nolint
# triangle_to_nowcast = triangle, #nolint
# delay_df = delay_df, #nolint
# disp_params = disp_params #nolint
# ) #nolint
```

In the above, we might want to make sure we include the necessary metadata if
the user passes it in, so we could write methods that dispatch on objects of
a particular type so that they spit back out things with dates if those were
inputted.
