---
title: "Example nowcast performance evaluation for model specification"
description: "An example of how to evaluate nowcast performance under different model specifications"
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Example nowcast performance evaluation for model specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette we demonstrate how to evaluate the performance of your `baselinenowcast` nowcast across multiple nowcast dates using visual checks and out of sample scoring using proper scoring rules fom the [`scoringutils`](https://epiforecasts.io/scoringutils/) package.
We'll start by doing this for a single model specification, and then show how you can use the results from a quantitative evaluation to help choose a `baselinenowcast` model specification.
Here we focus on optimising the training volume specification.
Before working through this vignette, we advise you to read both the Getting Started vignette and the NSSP nowcast vignette, both of which will demostrate how to produce a nowcast for a single nowcast date.


## Packages

We use the `baselinenowcast` package, `dplyr` for data-wrangling, `lubridate` for handling dates, and the `scoringutils` to evaluate out of sample nowcast performance.

```{r setup2, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(baselinenowcast)
library(lubridate)
library(dplyr)
library(scoringutils)
```

## Data

For this example, we will use the NSSP data, `syn_nssp_df` which is a synthetic dataset containing the number of incident cases indexed by reference and report date, both indexed at the daily scale.
The pathogen is not specified, but presumably this could represent the cases of a common respiratory syndrome such as influenza or COVID-19.
The data spans a typical respiratory virus season from 2025-2026, with reference dates ranging from October 25th, 2025 to May 5th, 2026, with reports ranging from October 25th, 2025 to August 23rd, 2027.

```{r}
head(syn_nssp_df)
```

## Evaluating across a range of nowcast dates

We'll start by specifying the range of nowcast dates that we want to produce retrospective nowcasts for.
Since this data has no historical data before October 25th, 2025, we will start from December 10th, 2025 to April 29th, 2025, producing nowcasts each week on Wednesdays.
```{r}
nowcast_dates <- seq(
  from = ymd("2025-12-10"),
  to = ymd("2026-04-01"), by = "week"
)
```

### Specify a single model

We'll start by evaluating a single `baselinenowcast` specification.
We will set the maximum delay to 25 days, and then specify the total number of reference dates used for training the model as 3x the maximum delay using the `scale_factor` argument, and specify that half of the training data be used for delay estimation with the `prop_delay` argument.
Note that the maximum delay should be specified based on both the delay distribution (e.g. you can use `truncate_to_quantile()` to truncate to a delay which captures a specified percentile of the reports) and your own knowledge of the data generating process (e.g. delays of more than a few weeks might be expected to be data entry errors).

Later in this vignette, we will demonstrate how to use quantitative nowcast performance to evaluate the performance of different training volume specifications.
```{r}
max_delay <- 25
scale_factor <- 2.2
prop_delay <- 0.5
```

### Looping over multiple nowcast dates

Next we will set up a for loop to retrospectively produce nowcasts using the data that would have been available at each of the retrospective nowcast dates.
Note, running this loop will print out a number of informational messages at each iteration, informing you that the reporting triangle has been truncated to the maximum delay and telling you exactly what proportion of the training volume was use for delay estimation.
```{r}
# Create the empty dataframes to store the nowcasts
mult_nowcast_quantiles <- c()
mult_nowcast_draws <- c()

for (i in seq_along(nowcast_dates)) {
  # Filter to only use data that would have been available as of the nowcast
  # date by removing all reports and reference dates
  # on or after the date of the nowcast
  training_data <- syn_nssp_df |>
    filter(
      report_date < nowcast_dates[i],
      reference_date < nowcast_dates[i]
    )

  # Save the initial reports used for comparison
  initial_count_data <- training_data |>
    group_by(reference_date) |>
    summarise(initial_count = sum(count))

  # Generate the nowcast
  rep_tri_full <- as_reporting_triangle(training_data)
  rep_tri <- truncate_to_delay(rep_tri_full, max_delay = max_delay)
  # Note, could also add another loop to iterate through different training
  # volumes and history for delay estimation
  nowcast_draws <- baselinenowcast(rep_tri,
    scale_factor = scale_factor,
    prop_delay = prop_delay,
    draws = 1000
  )
  # Could evaluate the draws directly using CRPS, or summarise into quantiles
  # and evaluate the quantiles. Will show both options.
  nowcast_quantiles <-
    nowcast_draws |>
    group_by(reference_date) |>
    summarise(
      `q_.50` = median(pred_count),
      `q_.25` = quantile(pred_count, 0.25),
      `q_.75` = quantile(pred_count, 0.75),
      `q_.05` = quantile(pred_count, 0.05),
      `q_.95` = quantile(pred_count, 0.95),
      `q_.025` = quantile(pred_count, 0.025),
      `q_.975` = quantile(pred_count, 0.975)
    ) |>
    # Add metadata for this iteration of the loop
    mutate(
      nowcast_date = nowcast_dates[i],
      scale_factor = scale_factor,
      prop_delay = prop_delay
    )


  nowcast_quantiles_w_data <- nowcast_quantiles |>
    left_join(initial_count_data,
      by = "reference_date"
    )
  nowcast_draws_w_data <- nowcast_draws |>
    left_join(initial_count_data,
      by = "reference_date"
    )

  # Combine the nowcasts from each iteration
  mult_nowcast_quantiles <- bind_rows(
    mult_nowcast_quantiles,
    nowcast_quantiles_w_data
  )
  mult_nowcast_draws <- bind_rows(
    mult_nowcast_draws,
    nowcast_draws_w_data
  )
}
```

Next, we'll want to join the evaluation data to the nowcasts combined with the initial counts. For the purpose of evaluation, we will want to only evaluate against the reports that occurred on or before the maximum delay, otherwise, older nowcasts have a longer period of evaluation data than more recent nowcasts.

```{r}
eval_data <- syn_nssp_df |>
  mutate(delay = as.integer(report_date - reference_date)) |>
  filter(delay <= max_delay) |>
  group_by(reference_date) |>
  summarise(final_count = sum(count))

# Left join the evaluation data to each of the dataframes of multiple nowcasts
mult_nowcast_quantiles_w_data <- mult_nowcast_quantiles |>
  left_join(eval_data, by = "reference_date")
mult_nowcast_draws_w_data <- mult_nowcast_draws |>
  left_join(eval_data, by = "reference_date")
```

