---
title: "Example nowcast performance evaluation for model specification"
description: "An example of how to evaluate nowcast performance under different model specifications"
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Example nowcast performance evaluation for model specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette we demonstrate how to evaluate the performance of your `baselinenowcast` nowcast across multiple nowcast dates using visual checks and out of sample scoring using proper scoring rules from the [`scoringutils`](https://epiforecasts.io/scoringutils/) package.
We'll start by doing this for a single model specification, and then show how you can use the results from a quantitative evaluation to help choose a `baselinenowcast` model specification.
Here we focus on optimising the training volume specification.
Before working through this vignette, we advise you to read both the Getting Started vignette and the NSSP nowcast vignette, both of which will demonstrate how to produce a nowcast for a single nowcast date.


## Packages

We use the `baselinenowcast` package, `dplyr` and `tidyr` for data-wrangling, `glue` for combining strings, `lubridate` for handling dates, and the `scoringutils` to evaluate out of sample nowcast performance.

```{r setup2, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(baselinenowcast)
library(lubridate)
library(tidyr)
library(dplyr)
library(glue)
library(scoringutils)
library(gganimate)
library(gifski)
```

## Data

For this example, we will use the NSSP data, `syn_nssp_df` which is a synthetic dataset containing the number of incident cases indexed by reference and report date, both indexed at the daily scale.
The pathogen is not specified, but presumably this could represent the cases of a common respiratory syndrome such as influenza or COVID-19.
The data spans a typical respiratory virus season from 2025-2026, with reference dates ranging from October 25th, 2025 to May 5th, 2026, with reports ranging from October 25th, 2025 to August 23rd, 2027.

```{r}
head(syn_nssp_df)
```

## Evaluating across a range of nowcast dates

We'll start by specifying the range of nowcast dates that we want to produce retrospective nowcasts for.
Since this data has no historical data before October 25th, 2025, we will start from December 10th, 2025 to April 29th, 2026, producing nowcasts each week on Wednesdays.
```{r}
nowcast_dates <- seq(
  from = ymd("2025-12-10"),
  to = ymd("2026-04-01"), by = "week"
)
```

#### Set the maximum delay

The maximum delay should be specified based on both the delay distribution and your own knowledge of the data generating process (e.g. delays of more than a few weeks might be expected to be data entry errors).
Here we will use a plot of the cumulative delay distribution to identify when the majority of our cases have been reported, noting that the lower our maximum delay is, the less historical data we will need to train the model.
In our case, since we don't have many years of historical data, we might wish to choose to keep this relatively low.
```{r}
avg_delays <- syn_nssp_df |>
  mutate(delay = as.integer(report_date - reference_date)) |>
  group_by(delay) |>
  summarise(pmf = sum(count) / sum(syn_nssp_df$count)) |>
  mutate(cdf = cumsum(pmf))

cdf_delay <- ggplot(avg_delays) +
  geom_line(aes(x = delay, y = cdf)) +
  geom_hline(aes(yintercept = 0.95), linetype = "dashed") +
  theme_bw() +
  xlab("Delay (days)") +
  ylab("Cumulative Distribution")
cdf_delay
```

Based on this plot, we will set our maximum delay to 25 days.

### Specify a single model

We'll start by evaluating a single `baselinenowcast` specification.
With the maximum delay set to 25 days, we will then specify the total number of reference dates used for training the model as 2.2x the maximum delay using the `scale_factor` argument, and specify that half of the training data be used for delay estimation with the `prop_delay` argument.

Later in this vignette, we will demonstrate how to use quantitative nowcast performance to evaluate the performance of different training volume specifications.
```{r}
max_delay <- 25
scale_factor <- 2.2
prop_delay <- 0.5
```

### Looping over multiple nowcast dates

First, we will write a function that runs the `baselinenowcast` workflow and saves the nowcasts along with the initial data and metadata about the model run (e.g. nowcast date and desired model specification variables).
```{r}
fit_model <- function(raw_data,
                      nowcast_date = nowcast_dates[i],
                      max_delay = max_delay,
                      scale_factor = scale_factor,
                      prop_delay = prop_delay,
                      output_type = "draws",
                      model_name = "baselinenowcast") {
  # Remove all cases on/after the nowcast date
  training_data <- raw_data |>
    filter(
      report_date < nowcast_date,
      reference_date < nowcast_date
    )

  # Save the initial reports used for comparison
  initial_count_data <- training_data |>
    group_by(reference_date) |>
    summarise(initial_count = sum(count)) |>
    filter(reference_date >= max(reference_date) - days(max_delay))

  # Generate the nowcast
  rep_tri_full <- as_reporting_triangle(training_data)
  rep_tri <- truncate_to_delay(rep_tri_full, max_delay = max_delay)
  # Note, could also add another loop to iterate through different training
  # volumes and history for delay estimation
  nowcast_draws <- baselinenowcast(rep_tri,
    scale_factor = scale_factor,
    prop_delay = prop_delay,
    draws = 1000
  ) |>
    mutate(
      nowcast_date = nowcast_date,
      scale_factor = scale_factor,
      prop_delay = prop_delay,
      model_name = "baselinenowcast"
    )

  if (output_type == "quantiles") {
    nowcast_quantiles <-
      nowcast_draws |>
      group_by(reference_date) |>
      summarise(
        `q_.50` = median(pred_count),
        `q_.25` = quantile(pred_count, 0.25),
        `q_.75` = quantile(pred_count, 0.75),
        `q_.05` = quantile(pred_count, 0.05),
        `q_.95` = quantile(pred_count, 0.95),
        `q_.025` = quantile(pred_count, 0.025),
        `q_.975` = quantile(pred_count, 0.975)
      ) |>
      # Add metadata for this iteration of the loop
      mutate(
        nowcast_date = nowcast_date,
        scale_factor = scale_factor,
        prop_delay = prop_delay
      )


    nowcasts_w_data <- nowcast_quantiles |>
      left_join(initial_count_data,
        by = "reference_date"
      )
  } else if (output_type == "draws") {
    nowcasts_w_data <- nowcast_draws |>
      left_join(initial_count_data,
        by = "reference_date"
      )
  }

  return(nowcasts_w_data)
}
```

Next we will set up a for loop to retrospectively produce nowcasts using the data that would have been available at each of the retrospective nowcast dates.
Note, running this loop will print out a number of informational messages at each iteration, informing you that the reporting triangle has been truncated to the maximum delay and telling you exactly what proportion of the training volume was use for delay estimation.
```{r}
# Create the empty dataframes to store the nowcasts
mult_nowcast_quantiles <- c()
mult_nowcast_draws <- c()

for (i in seq_along(nowcast_dates)) {
  # Filter to only use data that would have been available as of the nowcast
  # date by removing all reports and reference dates
  # on or after the date of the nowcast

  draws_w_data <- fit_model(
    raw_data = syn_nssp_df,
    nowcast_date = nowcast_dates[i],
    max_delay = max_delay,
    scale_factor = scale_factor,
    prop_delay = prop_delay,
    model_name = "baselinenowcast",
    output_type = "draws"
  )
  quantiles_w_data <- fit_model(
    raw_data = syn_nssp_df,
    nowcast_date = nowcast_dates[i],
    max_delay = max_delay,
    scale_factor = scale_factor,
    prop_delay = prop_delay,
    model_name = "baselinenowcast",
    output_type = "quantiles"
  )

  # Combine the nowcasts from each iteration
  mult_nowcast_quantiles <- bind_rows(
    mult_nowcast_quantiles,
    quantiles_w_data
  )
  mult_nowcast_draws <- bind_rows(
    mult_nowcast_draws,
    draws_w_data
  )
}
```

Next, we'll want to join the evaluation data to the nowcasts combined with the initial counts. For the purpose of evaluation, we will want to only evaluate against the reports that occurred on or before the maximum delay, otherwise, older nowcasts have a longer period of evaluation data than more recent nowcasts.

```{r}
eval_data <- syn_nssp_df |>
  mutate(delay = as.integer(report_date - reference_date)) |>
  filter(delay <= max_delay) |>
  group_by(reference_date) |>
  summarise(final_count = sum(count)) |>
  ungroup()

# Left join the evaluation data to each of the dataframes of multiple nowcasts
mult_nowcast_quantiles_w_data <- mult_nowcast_quantiles |>
  left_join(eval_data, by = "reference_date")
mult_nowcast_draws_w_data <- mult_nowcast_draws |>
  left_join(eval_data, by = "reference_date")
```

### Visual comparison of nowcasts

We'll start by visualising the nowcasts compared to both the initial data available without nowcast and the final data available retrospectively, using the dataframe containing the quantiled nowcasts.
To make a static image, we will start by showing a probabilistic nowcast (gray) every 4 weeks alongside the final data (red) and the initial data (pink).

```{r}
nowcast_dates_to_visualize <- seq(
  from = ymd("2025-12-10"),
  to = ymd("2026-04-01"), by = "4 weeks"
)

mult_nowcasts_filtered <- mult_nowcast_quantiles_w_data |>
  filter(nowcast_date %in% nowcast_dates_to_visualize)


p <- ggplot(mult_nowcasts_filtered) +
  geom_line(
    aes(
      x = reference_date, y = initial_count,
      group = nowcast_date,
      linetype = "Data as of nowcast date"
    ),
    color = "pink"
  ) +
  geom_vline(aes(
    xintercept = nowcast_date,
    linetype = "Date of nowcast"
  )) +
  geom_line(aes(x = reference_date, y = `q_.50`, group = nowcast_date),
    color = "gray"
  ) +
  geom_ribbon(aes(
    x = reference_date, ymin = `q_.025`, ymax = `q_.975`,
    group = nowcast_date
  ), fill = "gray", alpha = 0.25) +
  geom_ribbon(aes(
    x = reference_date, ymin = `q_.25`, ymax = `q_.75`,
    group = nowcast_date
  ), fill = "gray", alpha = 0.25) +
  geom_line(
    aes(
      x = reference_date, y = final_count,
      group = nowcast_date,
      linetype = "Final evaluation data"
    ),
    color = "red"
  ) +
  theme_bw() +
  scale_linetype_manual(
    name = "Observed data",
    values = c(
      "Final evaluation data" = "solid",
      "Data as of nowcast date" = "solid",
      "Date of nowcast" = "dashed"
    ),
    breaks = c(
      "Final evaluation data",
      "Data as of nowcast date",
      "Date of nowcast"
    ),
    guide = guide_legend(
      override.aes = list(
        color = c(
          "Final evaluation data" = "red",
          "Data as of nowcast date" = "pink",
          "Date of nowcast" = "black"
        ),
        linewidth = 1
      )
    )
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y"
  ) +
  xlab("Reference date") +
  ylab("Incident cases") +
  ggtitle("Visual nowcast comparison")
p
```


Optionally, we can make this into an animated figure, or a "flipbook" using `gganimate`.
This can be saved as a GIF with the command `anim_save()`.
```{r}
p_anim <- ggplot(mult_nowcast_quantiles_w_data) +
  # Final evaluation data — not grouped by nowcast_date so it stays fixed
  geom_line(
    aes(
      x = reference_date, y = final_count,
      linetype = "Final evaluation data"
    ),
    color = "red"
  ) +
  # Data as of nowcast date
  geom_line(
    aes(
      x = reference_date, y = initial_count,
      linetype = "Data as of nowcast date"
    ),
    color = "pink"
  ) +
  # Nowcast vertical line
  geom_vline(
    aes(
      xintercept = as.numeric(nowcast_date),
      linetype = "Date of nowcast"
    ),
    color = "black"
  ) +
  # Nowcast median + ribbons
  geom_line(aes(x = reference_date, y = `q_.50`),
    color = "gray"
  ) +
  geom_ribbon(aes(x = reference_date, ymin = `q_.025`, ymax = `q_.975`),
    fill = "gray", alpha = 0.25
  ) +
  geom_ribbon(aes(x = reference_date, ymin = `q_.25`, ymax = `q_.75`),
    fill = "gray", alpha = 0.25
  ) +
  theme_bw() +
  scale_linetype_manual(
    name = "Observed data",
    values = c(
      "Final evaluation data" = "solid",
      "Data as of nowcast date" = "solid",
      "Date of nowcast" = "dashed"
    ),
    breaks = c(
      "Final evaluation data",
      "Data as of nowcast date",
      "Date of nowcast"
    ),
    guide = guide_legend(
      override.aes = list(
        color = c(
          "Final evaluation data" = "red",
          "Data as of nowcast date" = "pink",
          "Date of nowcast" = "black"
        ),
        linewidth = 1
      )
    )
  ) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  xlab("Reference date") +
  ylab("Incident cases") +
  # gganimate: one frame per nowcast_date, label in title
  labs(title = "Nowcast date: {frame_time}") +
  transition_time(nowcast_date) +
  ease_aes("linear")

animate(p_anim,
  nframes = n_distinct(mult_nowcast_quantiles_w_data$nowcast_date),
  fps = 1.5, width = 800, height = 500, renderer = gifski_renderer()
)

anim_save("nowcast_flipbook.gif")
```

### Evaluation of nowcast performance using scoringutils

To quantify nowcast performance, we can evaluate nowcasts using proper scoring rules from the `scoringutils` package.
Before proceeding, we strongly recommend reading the package documentation for `scoringutils` and walking through the various vignettes, all of which contain examples of evaluating forecasts for similar epidemiological datasets.
We will evaluate the nowcast draws using the continuous ranked probability score and the quantiles using the weighted interval score, both of which are proper scoring rules which approximate absolute error for probabilistic forecasts.
Additionally, we will compute the interval coverage which tells us about how well the model is calibrated by calculating the proportion of all observations which fall within a certain prediction interval, with a perfectly well-calibrated forecast having the exact same coverage as the nominal coverage e.g. 90% of observations fall within the 90% prediction interval.

We start by setting an evaluation window-- the days for which we want to evaluate.
Since all `max_delay` days are being nowcasted, we will evaluate all of them.
However, we might only want to evaluate a subset of these (e.g. the last 2 weeks).

```{r}
eval_window <- max_delay
draws_to_score <- mult_nowcast_draws_w_data |>
  filter(reference_date >= nowcast_date - days(eval_window))
quantiles_to_score <- mult_nowcast_quantiles_w_data |>
  filter(reference_date >= nowcast_date - days(eval_window))
```

We start by converting both of our dataframes containing multiple nowcast comparisons into `scoringutils` forecast objects.
For the probabilistic draws, we will convert to a forecast object using the `scoringutils` function `as_forecast_sample()`.

```{r}
su_draws <- as_forecast_sample(
  draws_to_score,
  forecast_unit = c(
    "nowcast_date",
    "reference_date",
    "scale_factor",
    "prop_delay"
  ),
  observed = "final_count",
  predicted = "pred_count",
  sample_id = "draw"
)
```

For the quantile dataframe we used for plotting, we first need to pivot it from wide to long and then we can pass it to `scoringutils`'s function `as_forecast_quantile()`
```{r}
quantiles_to_score_long <- quantiles_to_score |>
  pivot_longer(
    cols = starts_with("q_"),
    names_prefix = "q_",
    names_to = "quantile_level",
    values_to = "pred_count"
  ) |>
  mutate(quantile_level = as.numeric(quantile_level))

su_quantiles <- as_forecast_quantile(
  quantiles_to_score_long,
  forecast_unit = c(
    "nowcast_date",
    "reference_date",
    "scale_factor",
    "prop_delay"
  ),
  observed = "final_count",
  predicted = "pred_count",
  quantile_level = "quantile_level"
)
```

Next we will score both the samples and the quantiles using the `scoringutils` `score()` function.

```{r}
scores_quantiles <- score(su_quantiles)
scores_samples <- score(su_draws)

head(scores_quantiles)
head(scores_samples)
```

We now have a table of scores for both the draws and quantiles, which contain scores at every nowcast date and reference date.
To investigate these further, we might be interested in how nowcast performance varies across nowcast dates.
We can visualise this by first summarising the scores by nowcast date using the `summarise_scores()` function, and then plotting the result.
We'll focus on the quantile-based scores, looking both at WIS and the 50th and 90th prediction intervals.


```{r}
scores_by_nowcast_date <- scores_quantiles |>
  summarise_scores(by = "nowcast_date")

ggplot(scores_by_nowcast_date) +
  geom_line(aes(x = nowcast_date, y = wis)) +
  geom_point(aes(x = nowcast_date, y = wis)) +
  theme_bw() +
  xlab("Nowcast date") +
  ylab("Average WIS")

ggplot(scores_by_nowcast_date) +
  geom_line(aes(x = nowcast_date, y = interval_coverage_50),
    color = "darkblue"
  ) +
  geom_point(aes(x = nowcast_date, y = interval_coverage_50),
    color = "darkblue"
  ) +
  geom_hline(aes(yintercept = .50),
    linetype = "dashed",
    color = "darkblue"
  ) +
  geom_line(aes(x = nowcast_date, y = interval_coverage_90),
    color = "maroon4"
  ) +
  geom_point(aes(x = nowcast_date, y = interval_coverage_90),
    color = "maroon4"
  ) +
  geom_hline(aes(yintercept = .90),
    linetype = "dashed",
    color = "maroon4"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  xlab("Nowcast date") +
  ylab("Interval Coverage")
```

#### Computing average interval coverage

We can also compute interval coverage across the whole time period using the same `summarise_scores()` function.
```{r}
scores_summary <- scores_quantiles |>
  mutate(model = "baselinenowcast") |>
  summarise_scores(by = "model") |>
  select(model, interval_coverage_50, interval_coverage_90) |>
  rename(
    "Model" = model,
    "50% coverage" = interval_coverage_50,
    "90% coverage" = interval_coverage_90
  ) |>
  knitr::kable(digits = 3)
scores_summary
```

## Choosing a model specification using quantitative performance evaluation

So far, we have demonstrated how to visually and quantitatively evaluate a single nowcasting model, where we specified the training volume so that it was fairly close to the package defaults.
However, the decision of how to specify the model should not be done arbitrarily-- instead, we can use the retrospective nowcast performance to compare how well different specifications would have done, which can be used to inform our current specification.
In this example, we will just focus on varying the training volume specification, but the same process could be applied to many of the other options for specifying `baselinenowcast`, including but not limited to: accounting for weekday effects via stratification by weekday, sharing estimates across strata (e.g. letting the delay from all age groups be used to nowcast each individual age groups case counts), and changing the observation model.

Here, we will test out using larger proportions of the training data for delay estimation  (via the `prop_delay` argument to `baselinenowcast()`) and we will test out using more or less training volume (via the `scale_factor` argument to `baselinenowcast()`).

In this example, we are somewhat limited because we don't have that much training data, but you can expand these to vary more in your use-case. See `allocate_reference_times()` for more details on the constraints of the training volume.
```{r}
prop_delays <- c(0.5, 0.7)
scale_factors <- c(2.2, 2.3)

# Create the empty dataframes to store the nowcasts
mult_nowcast_quantiles <- c()
mult_nowcast_draws <- c()

for (i in seq_along(nowcast_dates)) {
  for (j in seq_along(prop_delays)) {
    for (k in seq_along(scale_factors)) {
      draws_w_data <- fit_model(
        raw_data = syn_nssp_df,
        nowcast_date = nowcast_dates[i],
        max_delay = max_delay,
        scale_factor = scale_factors[k],
        prop_delay = prop_delays[j],
        model_name = "baselinenowcast",
        output_type = "draws"
      )
      quantiles_w_data <- fit_model(
        raw_data = syn_nssp_df,
        nowcast_date = nowcast_dates[i],
        max_delay = max_delay,
        scale_factor = scale_factors[k],
        prop_delay = prop_delays[j],
        model_name = "baselinenowcast",
        output_type = "quantiles"
      )
      # Combine the nowcasts from each iteration
      mult_nowcast_quantiles <- bind_rows(
        mult_nowcast_quantiles,
        quantiles_w_data
      )
      mult_nowcast_draws <- bind_rows(
        mult_nowcast_draws,
        draws_w_data
      )
    }
  }
}
```

Join the evaluation data (same as before)
```{r}
mult_nowcast_quantiles_w_data <- mult_nowcast_quantiles |>
  left_join(eval_data, by = "reference_date")
mult_nowcast_draws_w_data <- mult_nowcast_draws |>
  left_join(eval_data, by = "reference_date")
```

Filter to evaluate only the portion that is being nowcasted.
```{r}
eval_window <- max_delay
draws_to_score <- mult_nowcast_draws_w_data |>
  filter(reference_date >= nowcast_date - days(eval_window))
quantiles_to_score <- mult_nowcast_quantiles_w_data |>
  filter(reference_date >= nowcast_date - days(eval_window))
```

We start by converting both of our dataframes containing multiple nowcast comparisons into `scoringutils` forecast objects.
For the probabilistic draws, we will convert to a forecast object using the `scoringutils` function `as_forecast_sample()`.

```{r}
su_draws <- as_forecast_sample(
  draws_to_score,
  forecast_unit = c(
    "nowcast_date",
    "reference_date",
    "scale_factor",
    "prop_delay"
  ),
  observed = "final_count",
  predicted = "pred_count",
  sample_id = "draw"
)
```

For the quantile dataframe we used for plotting, we first need to pivot it from wide to long and then we can pass it to `scoringutils`'s function `as_forecast_quantile()`
```{r}
quantiles_to_score_long <- quantiles_to_score |>
  pivot_longer(
    cols = starts_with("q_"),
    names_prefix = "q_",
    names_to = "quantile_level",
    values_to = "pred_count"
  ) |>
  mutate(quantile_level = as.numeric(quantile_level))

su_quantiles <- as_forecast_quantile(
  quantiles_to_score_long,
  forecast_unit = c(
    "nowcast_date",
    "reference_date",
    "scale_factor",
    "prop_delay"
  ),
  observed = "final_count",
  predicted = "pred_count",
  quantile_level = "quantile_level"
)
```

Next we will score both the samples and the quantiles using the `scoringutils` `score()` function.

```{r}
scores_quantiles <- score(su_quantiles)
scores_samples <- score(su_draws)

head(scores_quantiles)
head(scores_samples)
```

###  Summarise performance by model specification

```{r}
scores_summary <- scores_quantiles |>
  mutate(model = "baselinenowcast") |>
  summarise_scores(by = c("model", "prop_delay", "scale_factor")) |>
  select(
    model, prop_delay, scale_factor, wis, underprediction, overprediction, dispersion,
    interval_coverage_50, interval_coverage_90
  )

scores_summary_table <- scores_summary |>
  rename(
    "Model" = model,
    "Proportion for delay" = prop_delay,
    "Scale factor" = scale_factor,
    "50% coverage" = interval_coverage_50,
    "90% coverage" = interval_coverage_90,
    "WIS" = wis
  ) |>
  knitr::kable(digits = 3)
scores_summary
```
Make a bar chart to visualise the scores, broken down by underprediction, overprediction, and dispersion.
```{r}
scores_summary |>
  pivot_longer(cols = c("overprediction", "underprediction", "dispersion")) |>
  mutate(
    name = factor(name, levels = c(
      "overprediction",
      "dispersion",
      "underprediction"
    ))
  ) |>
  mutate(model_name = glue("pd:{prop_delay} sf:{scale_factor}")) |>
  ggplot() +
  geom_bar(aes(x = model_name, y = value, fill = name),
    stat = "identity", position = "stack"
  ) +
  theme_bw() +
  scale_fill_manual(
    name = "WIS component",
    values = c("maroon4", "blue4", "orange4")
  ) +
  theme(
    axis.text.x = element_text(
      vjust = 1,
      hjust = 1,
      angle = 45,
      size = 11
    )
  ) +
  xlab("Model specification") +
  ylab("WIS")
```
Lastly, compare interval coverage for each specification.
```{r}
scores_summary |>
  mutate(interval_coverage_90 = interval_coverage_90 - interval_coverage_50) |>
  pivot_longer(cols = c("interval_coverage_90", "interval_coverage_50")) |>
  mutate(
    model_name = glue("pd:{prop_delay} sf:{scale_factor}"),
    name = factor(name, levels = c(
      "interval_coverage_90",
      "interval_coverage_50"
    ))
  ) |>
  ggplot() +
  geom_bar(aes(x = model_name, y = value, alpha = name),
    stat = "identity", position = "stack"
  ) +
  theme_bw() +
  scale_alpha_manual(
    name = "Interval coverage",
    values = c(
      interval_coverage_90 = 0.9,
      interval_coverage_50 = 0.5
    ),
    labels = c(
      interval_coverage_90 = "90%",
      interval_coverage_50 = "50%"
    )
  ) +
  geom_hline(aes(yintercept = 0.5), linetype = "dashed") +
  geom_hline(aes(yintercept = 0.9), linetype = "dashed") +
  theme(
    axis.text.x = element_text(
      vjust = 1,
      hjust = 1,
      angle = 45,
      size = 11
    )
  ) +
  xlab("Model specification") +
  ylab("Interval coverage")
```
