---
title: "Example nowcast performance evaluation for model specification"
description: "An example of how to evaluate nowcast performance under different model specifications"
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Example nowcast performance evaluation for model specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate how to evaluate the performance of your `baselinenowcast` nowcast across multiple nowcast dates.
We'll use visual checks and out-of-sample scoring with proper scoring rules from the [`scoringutils`](https://epiforecasts.io/scoringutils/) package.
For more resources on forecast evaluation, we recommend checking out the `scoringutils` documentation describing various [scoring rules](https://epiforecasts.io/scoringutils/articles/scoring-rules.html) used for evaluating point and probabilistic forecasts.
We also recommend checking out materials from the [Nowcasting and Forecasting of Infectious Disease Dynamics (NFIDD)](https://nfidd.github.io/nfidd/sessions/forecast-evaluation.html) short course on forecast evaluation, which walks through many of the principles of forecast evaluation alongside worked examples.
Additionally, we evaluate the performance of `baselinenowcast` under different model specifications and compared to other nowcasting methods in our [publication](https://wellcomeopenresearch.org/articles/10-614).
We'll start by evaluating a single model specification, and then show how to compare multiple specifications to inform your modelling choices, specifically focusing on evaluating the performance under different training volume specifications.
Before working through this vignette, we advise you to read both the [Getting Started](baselinenowcast.html) and the [NSSP nowcast](nssp_nowcast.html) vignettes, both of which will demonstrate how to produce a nowcast for a single nowcast date.


## Packages

We use the `baselinenowcast` package, `dplyr` and `tidyr` for data-wrangling, `glue` for combining strings, `lubridate` for handling dates, `purrr` for efficient iterating, `ggplot2` and `gganimate` for static and animated plotting (respectively), and the `scoringutils` to evaluate out of sample nowcast performance.

```{r setup2, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(baselinenowcast)
library(lubridate)
library(tidyr)
library(dplyr)
library(glue)
library(purrr)
library(gganimate)
library(ggplot2)
library(scoringutils)
```

## Data

For this example, we will use the NSSP data, `syn_nssp_df` which is a synthetic dataset containing the number of incident cases indexed by reference and report date, both indexed at the daily scale.
The pathogen is not specified, but presumably this could represent the cases of a common respiratory syndrome such as influenza or COVID-19.
The data spans a typical respiratory virus season from 2025-2026, with reference dates ranging from October 25th, 2025 to May 5th, 2026, with reports ranging from October 25th, 2025 to August 23rd, 2027.

```{r}
head(syn_nssp_df)
```

## Evaluating across a range of nowcast dates

We'll start by specifying the range of nowcast dates that we want to produce retrospective nowcasts for.
Since this data has no historical data before October 25th, 2025, we will produce nowcasts each week on Wednesday from December 10th, 2025 to April 1st, 2026.
```{r}
nowcast_dates <- seq(
  from = ymd("2025-12-10"),
  to = ymd("2026-04-01"), by = "week"
)
```

#### Set the maximum delay

The maximum delay should be specified based on both the delay distribution and your own knowledge of the data generating process (e.g. delays of more than a few weeks might be expected to be data entry errors).
Here we will use a plot of the cumulative delay distribution to identify when the majority of our cases have been reported, noting that the lower our maximum delay is, the less historical data we will need to train the model.
Because of right-truncation in recent counts, we expect to see in the recent reference dates only the shorter delays, and depending on the size of your data, this could bias the overall delay distribution.
Alternatively, the `baselinenowcast` package provides a helper function to automatically determine an appropriate maximum delay based on a user-specified proportion of cases, see `?truncate_to_quantile` for more details.
In our case, since we don't have many years of historical data, we might wish to choose to keep this relatively low.
```{r}
avg_delays <- syn_nssp_df |>
  mutate(delay = as.integer(report_date - reference_date)) |>
  group_by(delay) |>
  summarise(pmf = sum(count) / sum(syn_nssp_df$count)) |>
  mutate(cdf = cumsum(pmf))

cdf_delay <- ggplot(avg_delays) +
  geom_line(aes(x = delay, y = cdf)) +
  geom_vline(aes(xintercept = 25), linetype = "dashed") +
  geom_hline(aes(yintercept = 0.95), linetype = "dashed") +
  theme_bw() +
  xlab("Delay (days)") +
  ylab("Cumulative Distribution")
cdf_delay
```

Because we observe that 95% of cases are reported within 25 days, we will set our maximum delay to 25 days.

### Specify a single model

We'll start by evaluating a single `baselinenowcast` specification.
With the maximum delay set to 25 days, we will specify the total number of reference dates used for training as 2.2 times the maximum delay using the `scale_factor` argument.
We'll also specify that half of the training data be used for delay estimation with the `prop_delay` argument.
```{r}
max_delay <- 25
scale_factor <- 2.2
prop_delay <- 0.5
```

### Looping over multiple nowcast dates

First, we will write a function that extracts the training data for each nowcast date, by excluding all cases indexed or reported before the nowcast date.
```{r}
get_training_data <- function(raw_data,
                              nowcast_date,
                              max_delay) {
  training_data <- filter(
    raw_data,
    report_date < nowcast_date,
    reference_date < nowcast_date
  )
  return(training_data)
}
```


Next we will set up a for loop to retrospectively produce nowcasts using the data that would have been available at each of the retrospective nowcast dates.
Note, running this loop will print out a number of informational messages at each iteration, informing you that the reporting triangle has been truncated to the maximum delay and telling you exactly what proportion of the training volume was used for delay estimation.

```{r, message = FALSE}
mult_nowcast_draws <- map_dfr(
  setNames(nowcast_dates, nowcast_dates),
  \(nowcast_date) {
    training_data <- get_training_data(
      raw_data = syn_nssp_df,
      nowcast_date = nowcast_date
    )

    initial_count_data <- training_data |>
      group_by(reference_date) |>
      summarise(initial_count = sum(count)) |>
      filter(reference_date >= max(reference_date) - days(max_delay))

    rep_tri_full <- as_reporting_triangle(training_data)
    rep_tri <- truncate_to_delay(rep_tri_full, max_delay = max_delay)

    baselinenowcast(rep_tri,
      scale_factor = scale_factor,
      prop_delay = prop_delay,
      draws = 1000
    ) |>
      left_join(initial_count_data, by = "reference_date")
  },
  .id = "nowcast_date",
  .progress = TRUE
)
```

Next, we'll create the data to evaluate against using the latest observations.
We'll only include reports that occurred within the maximum delay window to ensure a fair comparison across all nowcast dates, as including all reports would mean that earlier nowcast dates are compared against longer delays than more recent nowcast dates.
See our [publication](https://wellcomeopenresearch.org/articles/10-614) and [Wolffram et al.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011394) for further details.

```{r}
eval_data <- syn_nssp_df |>
  mutate(delay = as.integer(report_date - reference_date)) |>
  filter(delay <= max_delay) |>
  group_by(reference_date) |>
  summarise(final_count = sum(count)) |>
  ungroup()
```



### Visual comparison of nowcasts

To make a static image, we will start by visualising the nowcasts compared to both the initial data available without nowcasting and the final data available retrospectively, using the dataframe containing the nowcast quantiles.

Compute quantiles for visualisation.
```{r}
mult_nowcasts <- mult_nowcast_draws |>
  mutate(nowcast_date = as.Date(nowcast_date)) |>
  group_by(reference_date, nowcast_date) |>
  summarise(
    median = median(pred_count),
    q50th_lb = quantile(pred_count, 0.25),
    q50th_ub = quantile(pred_count, 0.75),
    q95th_lb = quantile(pred_count, 0.025),
    q95th_ub = quantile(pred_count, 0.975)
  ) |>
  left_join(mult_nowcasts_filtered |>
    select(nowcast_date, reference_date, initial_count) |>
    distinct(), by = c("nowcast_date", "reference_date")) |>
  left_join(eval_data, by = c("reference_date"))
```

We'll start by making a shared plotting function, which expects a dataframe with columns for the quantiled probabilistic nowcasts (gray), the final data (red), and the initial data (pink).
```{r}
plot_multiple_nowcasts <- function(data) {
  p <- ggplot(data) +
    geom_line(
      aes(
        x = reference_date, y = initial_count,
        group = nowcast_date,
        color = "Data as of nowcast date"
      )
    ) +
    geom_vline(aes(
      xintercept = nowcast_date,
      color = "Date of nowcast"
    ), linetype = "dashed") +
    geom_line(aes(x = reference_date, y = median, group = nowcast_date),
      color = "gray"
    ) +
    geom_ribbon(aes(
      x = reference_date, ymin = q50th_lb, ymax = q50th_ub,
      group = nowcast_date
    ), fill = "gray", alpha = 0.25) +
    geom_ribbon(aes(
      x = reference_date, ymin = q95th_lb, ymax = q95th_ub,
      group = nowcast_date
    ), fill = "gray", alpha = 0.25) +
    geom_line(
      aes(
        x = reference_date, y = final_count,
        group = nowcast_date,
        color = "Final evaluation data"
      )
    ) +
    scale_color_manual(
      name = "Observed data",
      values = c(
        "Final evaluation data" = "red",
        "Data as of nowcast date" = "pink",
        "Date of nowcast" = "black"
      )
    ) +
    scale_x_date(
      date_breaks = "1 month",
      date_labels = "%b %Y"
    ) +
    theme_bw() +
    xlab("Reference date") +
    ylab("Incident cases") +
    ggtitle("Visual nowcast comparison")
  return(p)
}
```


To make a static image, we will start by showing a probabilistic nowcast every 4 weeks alongside the final data (red) and the initial data (pink).
First we will filter the nowcast dates to once every four weeks.
```{r}
nowcast_dates_to_visualize <- seq(
  from = ymd("2025-12-10"),
  to = ymd("2026-04-01"), by = "4 weeks"
)

mult_nowcasts_filtered <- mult_nowcasts |>
  filter(nowcast_date %in% nowcast_dates_to_visualize)

p <- plot_multiple_nowcasts(data = mult_nowcasts_filtered)
p
```


However, we might be interested in isolating the performance of each individual weekly nowcast.
<!--- TODO: This doesn't work with a nowcast each week, and I am not sure why but need to troubleshoot --->
To achieve this, we can make this into an animated figure, or a "flipbook" using `gganimate`, where each slide is one nowcast.
```{r}
p_anim <- p +
  labs(title = "Nowcast date: {closest_state}") +
  transition_states(nowcast_date, transition_length = 0, state_length = 1) +
  ease_aes("linear")

animate(p_anim,
  nframes = n_distinct(mult_nowcasts$nowcast_date),
  fps = 2,
  width = 8,
  height = 5,
  units = "in",
  res = 100,
  renderer = magick_renderer()
)
```

This can be saved as a GIF with the command `anim_save()`.
```{r}
anim_save("nowcast_flipbook.gif",
  animation = p_anim
)
```

### Evaluation of nowcast performance using scoringutils

In addition to visually assessing nowcast performance, quantitative performance evaluation can help us rigorously test the performance of different models and assess the reliability and accuracy of the nowcasts produced.
To quantify nowcast performance, we can evaluate nowcasts using proper scoring rules from the `scoringutils` package.
Before proceeding, we strongly recommend reading the package documentation for `scoringutils` and walking through the vignette on [scoring rules](https://epiforecasts.io/scoringutils/articles/scoring-rules.html).
We also recommend checking out materials from the [Nowcasting and Forecasting of Infectious Disease Dynamics (NFIDD)](https://nfidd.github.io/nfidd/sessions/forecast-evaluation.html) short course on forecast evaluation, which walks through many of the principles of quantitative forecast evaluation alongside worked examples.

In this vignette, we will evaluate the quantile-based nowcasts using the weighted interval score (WIS), which is a proper scoring rule that approximates the [continuous ranked probability score (CRPS)](https://epiforecasts.io/scoringutils/articles/scoring-rules.html#crps) which is a probabilistic scoring rule for sample-based forecasts that is a generation of the absolute error for probabilistic forecasts.

Additionally, we will compute the interval coverage, which tells us how well the model is calibrated by calculating the proportion of observations that fall within each prediction interval.
A perfectly well-calibrated forecast would have the exact same coverage as the nominal coverage (e.g., 90% of observations fall within the 90% prediction interval).

Note: The `scoringutils` package also supports sample-based scoring using the continuous ranked probability score (CRPS) if you prefer to work with draws rather than quantiles.
For this vignette, we focus on WIS and its components:
- **Overprediction**: penalty for predicting too high
- **Underprediction**: penalty for predicting too low
- **Dispersion**: penalty for prediction interval width

We'll use the `as_forecast_sample()` function in `baselinenowcast`, which ingests a `baselinenowcast_df` object, the evaluation data (a dataframe of the total cases observed within the `max_delay` for each reference date), and the `forecast_unit`, the unit on which to score the nowcasts, and returns a `scoringutils` `forecast_sample` object.
See `?as_forecast_sample` for more details.
We'll then using the `scoringutils` `as_forecast_quantile()` function to convert from samples to quantiles.
```{r}
su_draws <- as_forecast_sample(
  mult_nowcast_draws,
  latest_obs = eval_data,
  observed = "final_count",
  forecast_unit = c("reference_date", "nowcast_date", "model")
)

su_quantiles <- as_forecast_quantile(su_draws)
```

Now we'll call the `scoringutils` `score()` function to score the quantile-based nowcasts.
```{r}
scores_quantiles <- score(su_quantiles)

head(scores_quantiles)
```

We now have a table of scores of WIS and its components at every nowcast date and reference date.
Let's first summarise the scores overall using the `summarise_scores()` function from `scoringutils`.
```{r}
scores_summary <- scores_quantiles |>
  summarise_scores()
head(scores_summary)
```
This summary tells us that the model is overall biased slightly downward (indicated by the negative bias).
It slightly overcovers at 50% prediction intervals (meaning the prediction intervals are a bit too wide) but is very well-calibrated at 90% prediction intervals (89.99% of observations fall within the 90% prediction intervals).
In terms of overall performance, the nowcasts are penalised more heavily for underprediction than overprediction (which corroborates the negative bias) and is most penalised for being over-dispersed (prediction intervals are too wide, as corroborated by the 50% interval coverage value).



To investigate these further, we can look at how nowcast performance varies across nowcast dates.
We'll visualize this by first summarizing the scores by nowcast date using the `summarise_scores()` function.
```{r}
scores_by_nowcast_date <- summarise_scores(scores_quantiles,
  by = "nowcast_date"
) |> mutate(nowcast_date = as.Date(nowcast_date))

ggplot(scores_by_nowcast_date) +
  geom_line(aes(x = nowcast_date, y = wis)) +
  geom_point(aes(x = nowcast_date, y = wis)) +
  theme_bw() +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y"
  ) +
  xlab("Nowcast date") +
  ylab("Average WIS")
```

While this tells us how the nowcast performance varies over time, it can be a bit difficult to interpret on its own because WIS scales with the counts of the observations, thus, we expect higher WIS when case counts are higher.

Another visual check we can do across nowcast dates is to look at the 50th and 90th prediction intervals by nowcast date compared to their nominal values (indicated by the dashed line), as ideally the model is equally well-calibrated across time and as counts increase and decrease.
```{r}
scores_long <- scores_by_nowcast_date |>
  pivot_longer(
    cols = starts_with("interval_coverage_"),
    names_to = "interval",
    names_prefix = "interval_coverage_",
    values_to = "interval_coverage_value"
  )

ggplot(scores_long) +
  geom_line(aes(
    x = nowcast_date, y = interval_coverage_value,
    color = interval
  )) +
  geom_point(aes(
    x = nowcast_date, y = interval_coverage_value,
    color = interval
  )) +
  geom_hline(aes(yintercept = 0.50),
    linetype = "dashed",
    color = "darkblue"
  ) +
  geom_hline(aes(yintercept = 0.90),
    linetype = "dashed",
    color = "maroon4"
  ) +
  scale_color_manual(
    name = "Prediction interval (%)",
    values = c(
      "50" = "darkblue",
      "90" = "maroon4"
    )
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y"
  ) +
  xlab("Nowcast date") +
  ylab("Interval Coverage")
```

Here, we see relatively robust prediction interval coverage across the range of nowcast dates.

## Choosing a model specification using quantitative performance evaluation

So far, we have demonstrated how to visually and quantitatively evaluate a single nowcasting model, where we specified the training volume so that it was fairly close to the package defaults.
However, the decision of how to specify the model should not be done arbitrarily-- instead, we can use the retrospective nowcast performance to compare how well different specifications would have done, which can be used to inform our current specification.
In this example, we will just focus on varying the training volume specification, but the same process could be applied to many of the other options for specifying `baselinenowcast`, including but not limited to: accounting for weekday effects via stratification by weekday, sharing estimates across strata (e.g. letting the delay from all age groups be used to nowcast each individual age groups case counts), and changing the observation model.

Here, we will test out using larger proportions of the training data for delay estimation  (via the `prop_delay` argument to `baselinenowcast()`) and we will test out using more or less training volume (via the `scale_factor` argument to `baselinenowcast()`).

In this example, we are somewhat limited because we don't have that much training data, but you can expand these to vary more in your use-case. See `allocate_reference_times()` for more details on the constraints of the training volume.
```{r}
prop_delays <- c(0.5, 0.7)
scale_factors <- c(2.2, 2.3)

# Create empty dataframe to store the nowcasts
mult_nowcast_quantiles <- NULL

for (i in seq_along(nowcast_dates)) {
  for (j in seq_along(prop_delays)) {
    for (k in seq_along(scale_factors)) {
      # Fit model and get quantile summaries
      quantiles_w_data <- fit_model(
        raw_data = syn_nssp_df,
        nowcast_date = nowcast_dates[i],
        max_delay = max_delay,
        scale_factor = scale_factors[k],
        prop_delay = prop_delays[j]
      )

      # Combine the nowcasts from each iteration
      mult_nowcast_quantiles <- bind_rows(
        mult_nowcast_quantiles,
        quantiles_w_data
      )
    }
  }
}
```

Join the evaluation data (same as before).
```{r}
mult_nowcast_quantiles_w_data <- left_join(mult_nowcast_quantiles,
  eval_data,
  by = "reference_date"
)
```

Convert to scoringutils format and score (same process as the single specification section).
```{r}
eval_window <- max_delay

# Prepare data for scoring using our helper function
su_quantiles <- prepare_for_scoring(
  quantiles_df = mult_nowcast_quantiles_w_data,
  eval_window = eval_window
)

# Score the forecasts
scores_quantiles <- score(su_quantiles)

head(scores_quantiles)
```

###  Summarise performance by model specification

```{r}
scores_summary <- scores_quantiles |>
  mutate(model = "baselinenowcast") |>
  summarise_scores(by = c("model", "prop_delay", "scale_factor")) |>
  select(
    model, prop_delay, scale_factor, wis, underprediction, overprediction,
    dispersion, interval_coverage_50, interval_coverage_90
  )

scores_summary_table <- scores_summary |>
  # nolint start
  rename(
    "Model" = model,
    "Proportion for delay" = prop_delay,
    "Scale factor" = scale_factor,
    "50% coverage" = interval_coverage_50,
    "90% coverage" = interval_coverage_90,
    "WIS" = wis
  ) |>
  # nolint end
  knitr::kable(digits = 3)
scores_summary_table
```
Make a bar chart to visualize the scores, broken down by underprediction, overprediction, and dispersion.
```{r}
scores_summary |>
  pivot_longer(cols = c("overprediction", "underprediction", "dispersion")) |>
  mutate(
    name = factor(name,
      levels = c(
        "overprediction",
        "dispersion",
        "underprediction"
      )
    ),
    model_name = glue("pd:{prop_delay} sf:{scale_factor}")
  ) |>
  ggplot() +
  geom_bar(aes(x = model_name, y = value, fill = name),
    stat = "identity", position = "stack"
  ) +
  theme_bw() +
  scale_fill_manual(
    name = "WIS component",
    values = c("maroon4", "blue4", "orange4")
  ) +
  theme(
    axis.text.x = element_text(
      vjust = 1,
      hjust = 1,
      angle = 45,
      size = 11
    )
  ) +
  xlab("Model specification") +
  ylab("WIS")
```
Lastly, compare interval coverage for each specification.
```{r}
scores_summary |>
  mutate(interval_coverage_90 = interval_coverage_90 - interval_coverage_50) |>
  pivot_longer(cols = c("interval_coverage_90", "interval_coverage_50")) |>
  mutate(
    model_name = glue("pd:{prop_delay} sf:{scale_factor}"),
    name = factor(name, levels = c(
      "interval_coverage_90",
      "interval_coverage_50"
    ))
  ) |>
  ggplot() +
  geom_bar(aes(x = model_name, y = value, alpha = name),
    stat = "identity", position = "stack"
  ) +
  theme_bw() +
  scale_alpha_manual(
    name = "Interval coverage",
    values = c(
      interval_coverage_90 = 0.9,
      interval_coverage_50 = 0.5
    ),
    labels = c(
      interval_coverage_90 = "90%",
      interval_coverage_50 = "50%"
    )
  ) +
  geom_hline(aes(yintercept = 0.50), linetype = "dashed") +
  geom_hline(aes(yintercept = 0.90), linetype = "dashed") +
  theme(
    axis.text.x = element_text(
      vjust = 1,
      hjust = 1,
      angle = 45,
      size = 11
    )
  ) +
  xlab("Model specification") +
  ylab("Interval coverage")
```

### Interpreting the results

When comparing model specifications, consider the following:

**1. Weighted Interval Score (WIS)**: Lower is better. Look for meaningful differences (e.g., >10% reduction) rather than small variations that may be due to randomness. The WIS provides an overall measure of forecast accuracy that balances sharpness and calibration.

**2. WIS Components**:
   - High **underprediction** suggests the nowcasts are systematically too low, which could lead to underestimating disease burden
   - High **overprediction** suggests the nowcasts are systematically too high, which could lead to overestimating disease burden
   - High **dispersion** suggests prediction intervals are too wide

**3. Interval Coverage**:
   - Should be close to nominal levels (50% and 90%)
   - Consistently low coverage (e.g., 40% for 50% intervals) suggests overconfidence—intervals are too narrow
   - Consistently high coverage (e.g., 95% for 90% intervals) suggests underconfidence—intervals are too wide
   - Well-calibrated forecasts build trust and support better decision-making

**Selecting a model specification**: In this example, you would select the specification that minimizes WIS while maintaining well-calibrated interval coverage. If specifications have similar WIS, prefer the one with better calibration or simpler parameterization (lower `scale_factor` requires less training data, which may be beneficial when historical data is limited).

The retrospective evaluation approach demonstrated here can be applied to other `baselinenowcast` specifications beyond training volume, including weekday stratification, cross-stratum sharing, and observation model choices.
